---
title: "ESM 244 Lab 1"
author: "Roshni Katrak-Adefowora"
date: "1/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in SF trees data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for data wrangling & summary statistics using functions in the `dplyr` package.

Find the top 5 highest observations of trees by legal_status. Do some wrangling. Make a graph.

```{r}
top_5_status <- sf_trees %>% 
  count(legal_status) %>%  #recognizes groups (dont need group_by), finds counts (replaces n function), puts in table (replaces summarize)
  drop_na(legal_status) %>%  #remove rows that contain NA value
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% #move tree count to first column position
  slice_max(tree_count, n=5) #identify rows with highest values, keep only 5
```

Make a graphof those top 5 observations by legal status.

```{r}
ggplot(data=top_5_status, aes(x=fct_reorder(legal_status, tree_count), y=tree_count))+ #order bars by tree count
  geom_col()+
  labs(x="Legal Status", y="Tree Count")+
  coord_flip()+
  theme_minimal()
```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees.

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% #look within species variable for specific "Blackwood Acacia" string
  select(legal_status, date, latitude, longitude)

#preview map, but ggplot doesn't recognize lat/long as spatial coordinates
ggplot(data=blackwood_acacia, aes(x=longitude, y=latitude))+
  geom_point()
```

### tidyr::separate() and unite() functions

Useful for combining or separating columns

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep="::")
```

Example: tidyr::Unite()

```{r}
sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep="_cool!_")
```

### Make some actual maps of blackwood acacia trees in SF

Use `st_as_sf()` to convert lat & long to spatial coordinates.

```{r}
blackwood_acacia_sp <- blackwood_acacia %>% 
  drop_na(longitude, latitude) %>% 
  st_as_sf(coords = c("longitude", "latitude"))

#assign coordinate reference system
st_crs(blackwood_acacia_sp) = 4326

ggplot(data=blackwood_acacia_sp)+
  geom_sf(color="darkgreen")
```

Read in SF roads shapefile

```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

#st_transform(sf_map) run in console to see existing projection info

st_transform(sf_map, 4326) #use st_transform because already has existing CRS

#map of SF roads
ggplot(data=sf_map)+
  geom_sf()
```

Combine blackwood acacia tree observations and SF roads map

```{r}
ggplot()+
  geom_sf(data=sf_map, size=0.1, color="darkgray")+
  geom_sf(data=blackwood_acacia_sp, size=0.5, color="red")+
  theme_void()
```

Now an interactive map:

```{r}
tmap_mode("view")

tm_shape(blackwood_acacia_sp)+
  tm_dots()
```

